Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-05-16 paperDistance/paper.tex
remove four HOGGs; add four HOGGs

% # To-do list
% - We need a standard name for the \tgas--\tmass--\PanSTARRS\ Intersection.
% - zeroth draft / re-write the method section
% - add ORCIDs like Hogg's

@@

Here we deliver a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallaxes, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
HOGG:What is a photometric parallax? <Enter here>
Most stellar models used to generate photometric distances or photometric
parallaxes have been physical models, based on gravity, fluid mechanics,
radiative transfer, nuclear reactions, and atomic and molecular transitions.
Because there are small issues with each of these components, the models are inaccurate in detail, and
the physical models therefore produce distance estimates that are biased.
In addition, they build a long list of physical assumptions (about
nuclear reactions, convection, and thermal timescales, for example)
into the parallax estimates.
When using physical models, it is impossible to deliver photometric parallax estimates that involve minimal assumptions.

@@

So, returning the inference, we use Bayes theorem to turn the likelihood of the noisy data into a posterior believe over the true y value.
\begin{eqnarray}
p(y_{\true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{\true,n}, \sigma_n)\,p(y_{\true,n}) \\
\mathrm{where}\;p(y_n \given y_{\true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{\true,n}, \sigma_n) \\
\mathrm{and}\;p(y_{\true,n}) &=& \mathcal{N}(y_{\true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}
Since everything is Gaussian, this can be done analytically by multiplying Gaussians together. In Figure \ref{fig:toy} lower left panel, the blue points represent the posterior belief of the true value $p(y_{\true,n} \given y_n, \tilde{\sigma}_n)$ for the highlighted black points. The red points are the associated true y values.
In Figure \ref{fig:toy}, lower right panel, the blue points show the expectation value and $1\sigma$ uncertainty of the posterior belief for all 1024 $y_n$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance, $t^2$, and the measurement variance, $\sigma_n^2$.
The points are colored by their measurement variance $\sigma_n^2$, which shows that noisier measurements are more influenced by the prior and get pulled closer to the center of the MML distribution, our prior. In contrast, more certain measurements remain closer to their measured values, $y_n$. We now have better estimates of each true y value by inferring each true y value from the measured y value and using an estimate of the true distribution the measured y value was drawn from as the prior. Bringing the toy model back to our original problem, we can get better estimates of the true parallax for each star by inferring the true parallax from the measured parallax and using the noise-deconvolved \cmd, an estimate of the true \cmd, as our prior.

\section{Assumptions and methods}

@@

Transition clause/sentence. We obtained the photometric parallaxes in this project
by building a data-driven model of \cmd\ of the stars in the \tgas\ data set,
and using it as a prior pdf for Bayesian inference.
The posterior pdfs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
often introduce bias).
We have shown that, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well, by
showing that the posterior distance estimates to stars in stellar
clusters [HOGG: Is this too general?] are much more clustered than likelihood-based distance
estimates.

@@

Come back to the assumptions. What happens if we change or violate them?
Put in citations
Long discussion of the noise model. We took serious short-cuts. Call out to Boris.

\subsection{Proper Dust Corrections}
Within our probabilistic framework, we improperly handle dust. We are taking a point estimate of the probabilistic dust model at a quantile of our distance posterior, correcting our photometry using this point estimate, and then re-deriving the prior. A full, proper account would sample the dust model at distances sampled from the posterior, and propagate those samples back into the prior during the iterative process. The reason we don't do a full probabilistic treatment, is that this proper dust distribution would not be a Gaussian distribution, which \xd\ requires. We could calculate this more complex dust distribution and estimate it as a Gaussian distribution to feed to \xd, however, this is beyond the scope of this paper.

\subsection{Stationarity Assumption}
Stationarity assumes both the low and high signal to noise objects are being drawn from the same distribution, independent of noise. This isn't precisely true because most \gaia\ objects have similar noise $(\sim 3 \, \mathrm{mas})$, but their signal (parallax) depends on their distance. Nearby stars have higher signal (parallax) because they are closer and the nearby stars are mostly main sequence stars due to the IMF, timing, and malmquist bias. However, we can see bright giants to much larger distances, and therefore lower signal or parallax, leading to lower signal to noise. So the red giant branch is a population of relatively lower signal to noise objects compared with the main sequence. But this does not violate the stationarity assumption in detail, because the low and high signal to noise objects have to be drawn from the same distribution within some feature or patch of the space. This is all assuming the Gaia noise model is correct. If the uncertainties are over estimated then we could be over deconvolving (generating a tighter distribution than the truth) And in some extreme over estimate of uncertainties the lower signal to noise giants could be pulled to the main sequence. But with the quoted uncertainties this seems unlikely. As you move up the giant branch, and probe farther giants, the signal to noise will continue to diminish, but because the color uncertainties are not large, the S/N spread within a patch of the \cmd\ is well mixed. Possibly a metallicity trend with distance, placing more nearby (higher S/N) objects at slightly different positions on the CMD pulling lower S/N objects to that distribution for our prior. But if something has small error bars is good data is good data both for inference and generating the prior

For very rare kinds of stars. Inference with our prior is bad for underrepresented populations in the \cmd. For example, it appears there are only $2$ white dwarfs in \gaia\ DR1. With them being very underrepresented in the prior, the amplitudes of the gaussians representing them is very small, so the WDs are more likely to be pulled to mimic a much farther away MS star.

@@

Compare with literature stellar models

>>>>

% # To-do list
% - We need a standard name for the \tgas--\tmass--\PanSTARRS\ Intersection.
% - zeroth draft / re-write the method section

@@

Here we deliver a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallaxes, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
That is, this is a data-driven or purely empirical \emph{photometric
  parallax} method.
A photometric parallax is a parallax (or distance) estimate based on
the photometry (colors and apparent magnitude or magnitudes) of a
star.
Fundamentally, photometric parallaxes capitalize on the fact that the
apparent magnitude of a star is a strong function of distance, and
that the absolute magnitude is a strong function of color (or surface
temperature) and evolutionary stage.

Most stellar models used to generate photometric
parallaxes have been \emph{physical} models, based on gravity, fluid mechanics,
radiative transfer, nuclear reactions, and atomic and molecular transitions.
Because there are small issues with each of these components, the models are inaccurate in detail, and
the physical models therefore produce parallax (and hence distance) estimates that are biased.
In addition, they build a long list of physical assumptions (about
nuclear reactions, convection, and thermal timescales, for example)
into the parallax estimates.
When using physical models, it is impossible to deliver photometric parallax estimates that involve minimal assumptions.

@@

So, returning the inference, we use Bayes theorem to turn the likelihood of the noisy data into a posterior believe over the true y value.
\begin{eqnarray}
p(y_{\true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{\true,n}, \sigma_n)\,p(y_{\true,n}) \\
\mathrm{where}\;p(y_n \given y_{\true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{\true,n}, \sigma_n) \\
\mathrm{and}\;p(y_{\true,n}) &=& \mathcal{N}(y_{\true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}
Since everything is Gaussian, this can be done analytically by multiplying Gaussians together. In Figure \ref{fig:toy} lower left panel, the blue points represent the posterior belief of the true value $p(y_{\true,n} \given y_n, \tilde{\sigma}_n)$ for the highlighted black points. The red points are the associated true y values.
In Figure \ref{fig:toy}, lower right panel, the blue points show the expectation value and $1\sigma$ uncertainty of the posterior belief for all 1024 $y_n$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance, $t^2$, and the measurement variance, $\sigma_n^2$.
The points are colored by their measurement variance $\sigma_n^2$, which shows that noisier measurements are more influenced by the prior and get pulled closer to the center of the MML distribution, our prior. In contrast, more certain measurements remain closer to their measured values, $y_n$. We now have better estimates of each true y value by inferring each true y value from the measured y value and using an estimate of the true distribution the measured y value was drawn from as the prior.

In detail, the posterior points shown in the lower-right panel of
Figure~\ref{fig:toy} look more concentrated (vertically) than the true
values shown in the upper-left panel.
This is because the points in the lower-right (posterior) panel show
not typical values or typical draws from the posterior; they show the
modes (and, with error bars, the variances) of the posterior pdfs.
When an original data point has a large variance, the posterior ends
up close to the prior, which has a mode at the center line; the
posterior modes therefore concentrate more towards the center line
than any representative sampling.
A plot of posterior samplings (as opposed to a plot of posterior modes) would
show a scatter like that in the upper-left (truth) panel.

Bringing the toy model back to our original problem:
We can get better estimates of the true parallax for each star by inferring the true parallax from the measured parallax and using the noise-deconvolved \cmd, an estimate of the true \cmd, as our prior.

\section{Assumptions and methods}\label{sec:method}

@@

Transition clause/sentence. We obtained the photometric parallaxes in this project
by building a data-driven model of \cmd\ of the stars in the \tgas\ data set,
and using it as a prior pdf for Bayesian inference.
The posterior pdfs for distance that we obtain are, in general, much
narrower than the likelihood functions delivered by the
\gaia\ Mission, and therefore the distance estimates (or,
equivalently, parallax estimates) are much more precise.
It is not surprising that a Bayesian inference provides more
precise inferences than the likelihood function alone; Bayesian
inferences bring in new information that decrease variance (but
often introduce bias).
We have shown that, in addition to pure precision improvements, at
least some aspects of \emph{accuracy} have been improved as well, by
showing that the posterior distance estimates to stars in (at least one) stellar
cluster are much more clustered than likelihood-based distance
estimates.

@@

In Section~\ref{sec:method} we listed a set of assumptions upon which
the photometric parallax method in this project is based.
Here we return to these assumptions, and discuss them critically in more
detail.
\begin{description}
\item[stationarity] Stationarity assumes both the low and high
  signal to noise objects are being drawn from the same distribution,
  independent of noise. This isn't precisely true because most
  \gaia\ objects have similar noise $(\sim 3 \, \mathrm{mas})$, but
  their signal (parallax) depends on their distance. Nearby stars have
  higher signal (parallax) because they are closer and the nearby
  stars are mostly main sequence stars due to the IMF, timing, and
  malmquist bias. However, we can see bright giants to much larger
  distances, and therefore lower signal or parallax, leading to lower
  signal to noise. So the red giant branch is a population of
  relatively lower signal to noise objects compared with the main
  sequence. But this does not violate the stationarity assumption in
  detail, because the low and high signal to noise objects have to be
  drawn from the same distribution within some feature or patch of the
  space. This is all assuming the Gaia noise model is correct. If the
  uncertainties are over estimated then we could be over deconvolving
  (generating a tighter distribution than the truth) And in some
  extreme over estimate of uncertainties the lower signal to noise
  giants could be pulled to the main sequence. But with the quoted
  uncertainties this seems unlikely. As you move up the giant branch,
  and probe farther giants, the signal to noise will continue to
  diminish, but because the color uncertainties are not large, the S/N
  spread within a patch of the \cmd\ is well mixed. Possibly a
  metallicity trend with distance, placing more nearby (higher S/N)
  objects at slightly different positions on the CMD pulling lower S/N
  objects to that distribution for our prior. But if something has
  small error bars is good data is good data both for inference and
  generating the prior

  For very rare kinds of stars. Inference with our prior is bad for
  underrepresented populations in the \cmd. For example, it appears
  there are only $2$ white dwarfs in \gaia\ DR1. With them being very
  underrepresented in the prior, the amplitudes of the gaussians
  representing them is very small, so the WDs are more likely to be
  pulled to mimic a much farther away MS star.
\item[big data] HOGG 
\item[noise model] HOGG
\item[dust] Within our probabilistic framework, we improperly handle
  dust. We are taking a point estimate of the probabilistic dust model
  at a quantile of our distance posterior, correcting our photometry
  using this point estimate, and then re-deriving the prior. A full,
  proper account would sample the dust model at distances sampled from
  the posterior, and propagate those samples back into the prior
  during the iterative process. The reason we don't do a full
  probabilistic treatment, is that this proper dust distribution would
  not be a Gaussian distribution, which \xd\ requires. We could
  calculate this more complex dust distribution and estimate it as a
  Gaussian distribution to feed to \xd, however, this is beyond the
  scope of this paper.
\item[mixture of Gaussians] HOGG
\item[no physics] HOGG
\end{description}

@@

Anderson: Compare with literature stellar models?? This is an enhanced goal; strike this.
